Free surface computations in mantle convection models

### 2015-10-30 free-surface-paper.tex
part one

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{color, soul}

@@

\author{I. Rose, B. Buffett, and T. Heister}

@@

All of the approaches to free surface simulations have been subject to an instability which has been 
variously termed a ``sloshing,'' or ``druknen sailor'' instability \citep{kaus2010stabilization, duretz2011discretization, kramer2012implicit}. 
This instability, arising from the large density contrast typical at a free surface (compared with the much smaller 
internal density contrasts), severely limits the maximum stable timestep for free surface computations.
Frequently, the maximum stable timestep is several orders of magnitude smaller than that for an otherwise 
equivalent model with free-slip boundary conditions.

@@

Furthermore, since we are interested in solving these equations for a domain with a free surface, 
we need an equation to track that surface.
The Lagrangian equation for time evolution of the free surface can be written
\begin{equation}
\frac{d \zeta}{dt} = \mathbf{u \cdot n} \quad \textrm{on  }  \partial \Omega_\textrm{free surface}
\label{eq:surface_evolution}
\end{equation}
where $\zeta$ is the surface coordinate of the free surface, and $\partial \Omega_\textrm{free surface}$ is the 
part of the domain boundary corresponding to the free surface.

@@

In order to better understand the time evolution of Equations~\eqref{eq:momentum_mass_conservation}-\eqref{eq:surface_evolution} 
we will consider the eigenvalues of a simplified homogeneous system. By setting $\rho = \rho_0$, we obtain the system
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime = 0
\label{eq:homogeneous_stokes}
\end{equation}

@@

\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
+ \int_{\partial \Omega_N(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
\label{eq:weak_stokes}
\end{equation}

\begin{equation}
\int_{\Omega(t)} q \nabla \cdot \mathbf{u} = 0
\label{eq:weak_incompressible}
\end{equation}
where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
$\Omega$ and $\partial \Omega_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
Note that since the free surface can move, the shape of the domain is a function of time.
It is most convenient to integrate over the domain at the current timestep, which corresponds to an explicit scheme.
However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep
is much more stable, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.

The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.
Rather than analyzing finite deformation to the free surface (a nonlinear problem),
we will make the analytically useful approximation of small deformations about the hydrostatic 
reference surface and analyze their stability.
We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
represents the (small) topography of the free surface relative to that reference surface.
On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
The stress on this reference surface can be approximated by using the first order Taylor series
expansion of the hydrostatic pressure profile:
\begin{equation}
\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = \rho_0 g \zeta \mathbf{I}
\label{eq:hydrostatic}
\end{equation}
Equation~\eqref{eq:weak_stokes} then becomes

@@

\subsection{Remeshing}
In ALE calculations the internal mesh velicity is undetermined.
In general, one wants to smoothly deform the mesh so as to preserve its regularity, 
avoiding inverted or otherwise poorly conditioned cells.
The mesh deformation can be calcluated in many different ways, including algebraic \citep[e.g.][]{thieulot2011fantom} 
and PDE based approaches.

>>>>

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{color, soul}
\usepackage{url}

\usepackage{todonotes}
\newcommand\timo[1]{\todo[inline]{Timo: #1}}

@@

\author{I. Rose, B. Buffett, and Timo Heister\fnref{ref3}}

\fntext[ref3]{\url{heister@clemson.edu}, partially supported by the Computational Infrastructure in
Geodynamics initiative (CIG), through the National Science Foundation under Award No. EAR-0949446 
and The University of California—Davis, and National Science Foundation grant DMS1522191.}

@@

All of the approaches to free surface simulations have been subject to an instability which has been 
variously termed a ``sloshing,'' or ``drunken sailor'' instability \citep{kaus2010stabilization, duretz2011discretization, kramer2012implicit}. 
This instability, arising from the large density contrast typical at a free surface (compared with the much smaller 
internal density contrasts), severely limits the maximum stable timestep for free surface computations.
Frequently, the maximum stable timestep is several orders of magnitude smaller than that for an otherwise 
equivalent model with free-slip boundary conditions.

@@

Formally, this gives rise to the following time dependent, coupled system with unknowns $\mathbf{u}(t)$ and $\mathbf{p}(t)$:
\begin{equation}
\begin{aligned}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p &= \rho \mathbf{g} \qquad \text{in} \qquad \Omega(t)\\
\nabla \cdot \mathbf{u} &= 0  \qquad \text{in} \qquad  \Omega(t)
\end{aligned}
\label{eq:final_system}
\end{equation}
defined on the bounded, moving domain $\Omega(t)\subset \mathbb{R}^d$ with boundary $\partial \Omega = \Gamma_0 \cup \Gamma_F$
split into a fixed (Dirichlet) and free surface part $\Gamma_0$ and $\Gamma_F$, respectively.
The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by a mesh velocity $\mathbf{v}(t)$
\timo{Do we need $\Gamma_N$ here?}
\begin{equation}
 \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{v}(t) \text{ d} t.
\end{equation}
The mesh velocity in normal direction is given by the velocity $\mathbf{u}(t)$ on the free surface
\begin{equation}
 \mathbf{v}(t) \cdot n = \mathbf{u}(t) \cdot n \text{on} \Gamma_F
\end{equation}
and set to zero on the rest of the boundary. We defer the discussion of the velocity $\mathbf{v}$ in the interior of the domain
to Section~\ref{sec:remeshing}.
Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{v}(t) \text{ d} t$.

@@

In order to better understand the time evolution of the system in \eqref{eq:final_system}
we will consider the eigenvalues of a simplified homogeneous system. By setting $\rho = \rho_0$, we obtain the system
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime = 0
\label{eq:homogeneous_stokes}
\end{equation}

@@

\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
+ \int_{\Gamma_N(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
\label{eq:weak_stokes}
\end{equation}
\timo{change boundary $\Gamma_N$ with new notation above.}
\begin{equation}
\int_{\Omega(t)} q \nabla \cdot \mathbf{u} = 0
\label{eq:weak_incompressible}
\end{equation}
where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
$\Omega$ and $\partial \Gamma_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
Note that since the free surface can move, the shape of the domain is a function of time.
It is most convenient to integrate over the domain at the current timestep, which corresponds to an explicit scheme.
However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep
is much more stable, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.

\timo{TODO: update. write as solving on $\Omega(t^n)$ instead of $\Omega(t^{n+1})$}
The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.
Rather than analyzing finite deformation to the free surface (a nonlinear problem),
we will make the analytically useful approximation of small deformations about the hydrostatic 
reference surface and analyze their stability.
We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
represents the (small) topography of the free surface relative to that reference surface.
On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
The stress on this reference surface can be approximated by using the first order Taylor series
expansion of the hydrostatic pressure profile:
\begin{equation}
\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = \rho_0 g \zeta \mathbf{I}
\label{eq:hydrostatic}
\end{equation}
Equation~\eqref{eq:weak_stokes} then becomes

@@

\subsection{Remeshing}
\label{sec:remeshing}

In ALE calculations the internal mesh velicity is undetermined.
In general, one wants to smoothly deform the mesh so as to preserve its regularity, 
avoiding inverted or otherwise poorly conditioned cells.
The mesh deformation can be calcluated in many different ways, including algebraic \citep[e.g.][]{thieulot2011fantom} 
and PDE based approaches.
